Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-21 doc/mc-paper.tex
Polishing

We know of no other
work addressing the question of certifying the output of a knowledge
compiler.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  negation normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.

  For Boolean formula $\phi$, and ring weight function $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

@@

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1 \leq i \leq k} \phi_i, w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1 \leq i \leq k} \phi_i, w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
Every POG must have a designated {\em root node} $\noder$.

@@

The combination of these two optimization guarantees that 1) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and 2) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results (Appendix~\ref{app:experiments}) shows that these
optimizations yield substantial benefits.

@@

%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1 \leq i \leq k}
%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
%% the sum of $t_1, \ldots, t_k$.

>>>>

We know of no other
work on certifying the output of a knowledge compiler.

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

@@

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
Every POG has a designated {\em root node} $\noder$.

@@

The combination of these two optimization guarantees that i) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results (Appendix~\ref{app:experiments}) shows that these
optimizations yield substantial benefits.

@@

%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
%% the sum of $t_1, \ldots, t_k$.
